Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-04 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: added comment to ab; moved variable-p idea to conclusions

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) = \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
% \begin{equation}
% f = \exp(\varphi)\,,
% \end{equation}
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.

@@

In general \dots
%
\hs{Describe PN works.}

@@

In the ParSpec expansion this is accomplished by performing a bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, namely
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{n}$ and $\tau_{k}^{n}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs;
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on a given modified gravity theory is encapsulated in the
dimensionless parameter $\gamma$, given by
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on a parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

@@

% \hs{TODO: in figures and tables use $N_{\rm max} = 0$ ($N_{\rm max} = 1$) instead of $j=0$ ($j=0,1$).}

@@

We now describe the waveform model used in our paper to measure properties of a
BBH ringdown.
%
As in Refs.~\cite{Brito:2018rfr,Ghosh:2021mrv}, we use an IMR BBH waveform
model where the parameters describing the remnant object are left additionally
free and estimated directly from the data.

@@

The polarization of the GW signal (in the observer's frame) is,
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t)\,,
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical
harmonics described by standard angular dependence indices $l$ and $m$.
%
% --------------------------------
% HS: maybe not necessary anymore:
% \footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.}
% --------------------------------
%
As our baseline model, we use the computationally efficient (time-domain) multipolar waveform model for quasicircular spin-aligned BBH mergers described in Ref.~\cite{Mihaylov:2021bpf} (henceforth referred to as \SEOB{}\footnote{This waveform model is available in \texttt{LALSuite}~\cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA} waveform approximant.}) which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$~\cite{Cotesta:2018fcv,Mihaylov:2021bpf}.
%
The model uses an effective-one-body (EOB) approach combined with a
post-adiabatic solution of the equations of motion to describe the
inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$.
%
An accurate description of the merger is incorporated through calibration with
NR simulations, as described in Ref.~\cite{Cotesta:2018fcv}, along with
information for the merger and ringdown phases, from BH perturbation theory.
The merger-ringdown waveform, $h_{l m}^\mathrm{merger-RD}$, is then stitched to
inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) =
h_{l m}^\textrm{insp-plunge}\, \theta(t_\textrm{match}^{l m} - t)
+ h_{l m}^\textrm{merger-RD}\,\theta(t-t_\textrm{match}^{l m})\,,
\end{equation}
%
where $\theta(t)$ is the Heaviside step function.
%
The merger-ringdown waveform is expressed as an exponentially damped
sinusoid~\cite{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})}\,,
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} = {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})
= \omega_{l m 0} - i / \tau_{l m 0}\,,
\end{align}
%
are the complex frequencies of the fundamental ($n=0$ overtone) QNMs of the remnant BH.
%
The frequencies $\omega_{l m 0}$ and damping times $\tau_{l m 0}$ of the $(l,
m)$ QNM can be read off from the real and imaginary parts of $\sigma_{l m 0}$
respectively.
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined
in Ref.~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

In the $\SEOB$ model~\cite{Mihaylov:2021bpf}, the complex frequencies
$\sigma_{l m 0}$ are computed by first determining the final mass and spin from
estimates of the initial masses and spins through NR-fitting-formulas~\cite{Taracchini:2013rva,Hofmann:2016yih}
and then converting them to the complex frequencies using BH perturbation
theory-inspiral analytical fits~\cite{Berti:2005ys,Berti:2009kk}.
%
Hence,
%
\begin{subequations}
\begin{align}
\omega_{l m 0}^{\text{GR}} &= \omega_{l m 0}^{\text{GR}}(m_1, m_2, \chi_1, \chi_2)\,,\\
\tau _{l m 0}^{\text{GR}} &= \tau _{l m 0}^{\text{GR}}(m_1, m_2, \chi_1, \chi_2)\,.
\end{align}
\end{subequations}
%
where $(\omega_{l m 0}^{\text{GR}}, \tau_{l m 0}^{\text{GR}} )$ refer to the
QNM predictions in baseline \SEOB{} model.

@@

Following the ParSpec expansion, our QNM frequencies are expressed through functions,
%
\begin{subequations}
\begin{align}
\omega_k &= \omega_k(m_1, m_2, \chi_1, \chi_2,\ell, \{\delta \omega_k^{(j)}\}),\\
\tau_k   &= \tau _k(m_1, m_2, \chi_1, \chi_2, \ell, \{\delta \tau_k^{(j)}\}),
\end{align}
\end{subequations}
%
where we fixed a certain value of $p$, and the estimates of $(m_1,\, m_2,\, \chi_1,\, \chi_1)$ are used to predict
the final BH mass $M_f$ and spin $\chi_f$ through~\cite{Taracchini:2013rva,Hofmann:2016yih}.
%
% HS: we implicitly fixed p here. Could we leave it free in PE study. This is similar
% to the 'number of dimensions'-like studies. Here it would tells us quite generically
% what is the dimension of the modification to the EH-action we include!
%
Additionally, both $\omega_k$ and $\tau_k$ depend on the coupling constant $\ell$
defined in Eq.~\eqref{eq:def_gamma} as well as the sets, $\{\delta
\omega_k^{(n)}\},$ and $\{\delta \tau_k^{(n)}\}$, where $n$ ranges from zero to
the power of spin in Eqs.~\eqref{eq:kerr_expansion} up to which we include
corrections.
%
For example, if we restrict ourselves to corrections with $n=0$, or
$\chi_f^0$, to the $k=(2,2,0)$ mode only, then we would just have two
additional parameters, $\{\delta \omega_0^{(0)}, \delta \tau_0^{(0)}\}$.
%
In this paper, we will restrict ourselves to corrections to the $n=0$ and
$n=1$ terms, i.e., leading and next-to-leading order corrections to the spin,
$\chi_f$.

@@

>>>>

Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH
solutions.
%
In the first class, the first derivative of the coupling function $f'(\varphi) = \dd f  / \dd \varphi$
is always nonzero and BHs are known to always support scalar hair.
%
This is the case of Einstein-dilaton-Gauss-Bonnet (EdGB) gravity, for which $f(\varphi) = \exp(\varphi)$~\cite{Kanti:1995vq}.
%
% \begin{equation}
% f = \exp(\varphi)\,,
% \end{equation}
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, the theory admits the same stationary, asymptotically flat BH
solutions of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Macedo:2019sem,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the Gaussian $f(\varphi) \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} and
the quadratic $f(\varphi) \propto \varphi^2$~\cite{Silva:2017uqg} coupling functions.

@@

In general \dots
%
\hs{Alessandra, can you include a short description of~\cite{Sennett:2019bpc} for the quartic EFT?}

@@

In the ParSpec expansion this is accomplished by performing a bivariate
expansion of Eq.~\eqref{eq:general_deviation} in the detector-frame BH's mass
$M$ and spin $\chi$, namely
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{k} &= \frac{1}{M} \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \omega^{(n)}_{k} \, \left( 1 + \gamma \delta \omega^{(n)}_{k} \right), \\
\tau_{k}   &= M     \sum_{n = 0}^{N_{\rm max}} \, \chi^{n} \tau^{(n)}_{k}   \, \left( 1 + \gamma \delta \tau^{(n)}_{k} \right),
\end{align}
\end{subequations}
%
where $\omega_{k}^{n}$ and $\tau_{k}^{n}$ are dimensionless coefficients of the
expansion in spin for Kerr QNMs;
%
$\delta\omega_{k}^{(n)}$ and $\delta\tau_{k}^{(n)}$ are source-independent
dimensionless coefficients that characterize the corrections to the Kerr QNM at
each spin-order;
%
and finally $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence on a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, given by
%
\begin{equation}
\gamma
%
= \left(\frac{\ell}{M_{\rm s}}\right)^{p}
%
= \left[
\frac{\ell c^2 (1 + z)}{G M}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on a parameter $\ell$, with dimensions of length (and sets the
lengthscale at which beyond-GR modifications become important), and
the exponent $p$ which is related to how the beyond-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\ell$ dimensionless by using the other
lenghtscale associated to the remnant black hole, i.e., its source-frame mass
$M_{\rm s}$, which we can also write in terms of the detector-frame mass $M$ through
the redshift $z$~\cite{Krolak:1987ofj}.

@@

@@

We now describe the waveform model used in our paper to infer properties of a
BBH ringdown.
%
As in Refs.~\cite{Brito:2018rfr,Ghosh:2021mrv}, we use an IMR BBH waveform
model where the parameters describing the remnant object are left additionally
free and estimated directly from the data.

@@

The polarization of the GW signal (in the observer's frame) is,
%
\begin{equation}
h_+(\iota,\varphi_0;t ) - i h_\times(\iota,\varphi_0;t) = \sum_{l, m} {}_{-\!2}Y_{l m}(\iota,\varphi_0)\, h_{l m}(t)\,,
\end{equation}
%
where ${}_{-\!2}Y_{l m}(\iota,\varphi_0)$ are the $-2$ spin-weighted spherical
harmonics described by standard angular dependence indices $l$ and $m$.
%
% --------------------------------
% HS: maybe not necessary anymore:
% \footnote{Note that we use $l$ to denote the angular dependence index and $\ell$ to denote the coupling constant in Sec.~\ref{sec:review_parspec}.}
% --------------------------------
%
As our baseline model, that is, the GR model upon which non-GR modifications will be added to, we use the computationally efficient (time-domain) multipolar waveform model for quasicircular spin-aligned BBH mergers described in Ref.~\cite{Mihaylov:2021bpf} (henceforth referred to as \SEOB{}\footnote{This waveform model is available in \texttt{LALSuite}~\cite{lalsuite} as the \texttt{SEOBNRv4HM\_PA} waveform approximant.}) which contains the modes, $(l, |m|)=(2,2),(2,1)$, $(3,3)$, $(4,4)$, and $(5,5)$~\cite{Cotesta:2018fcv,Mihaylov:2021bpf}.
%
The model uses an effective-one-body (EOB) approach combined with a
post-adiabatic solution of the equations of motion to describe the
inspiral-plunge waveform, $h_{l m}^\mathrm{insp-plunge}$.
%
An accurate description of the merger is incorporated through calibration with
NR simulations, as described in Ref.~\cite{Cotesta:2018fcv}, along with
information for the merger and ringdown phases, from BH perturbation theory.
The merger-ringdown waveform, $h_{l m}^\textrm{merger-RD}$, is then stitched to
inspiral-plunge waveform, $h_{l m}^\textrm{insp-plunge}$ at a certain time $t = t^{\textrm{match}}_{l m}$, as
%
\begin{equation}
h_{l m}(t) =
h_{l m}^\textrm{insp-plunge}\, \theta(t_\textrm{match}^{l m} - t)
+ h_{l m}^\textrm{merger-RD}\,\theta(t-t_\textrm{match}^{l m})\,,
\end{equation}
%
where $\theta(t)$ is the Heaviside step function.
%
The merger-ringdown waveform is expressed as an exponentially damped
sinusoid~\cite{Bohe:2016gbl,Cotesta:2018fcv,Mihaylov:2021bpf}
%
\begin{equation}
\label{RD}
h_{l m}^{\textrm{merger-RD}}(t) = \nu \ \tilde{A}_{l m}(t)\ e^{i \tilde{\phi}_{l m}(t)} \ e^{-i \sigma_{l m 0}(t-t^{\textrm{match}}_{l m})}\,,
\end{equation}
%
where
%
\begin{align}
\sigma_{l m 0} = {\rm Re}(\sigma_{l m 0}) + i \, {\rm Im}(\sigma_{l m 0})
= \omega_{l m 0} - i / \tau_{l m 0}\,,
\end{align}
%
are the complex frequencies of the fundamental ($n=0$ overtone) QNMs of the remnant BH.
%
The frequencies $\omega_{l m 0}$ and damping times $\tau_{l m 0}$ of the $(l,
m)$ QNM can be read off from the real and imaginary parts of $\sigma_{l m 0}$
respectively.
%
The functions $\tilde{A}_{l m}(t)$ and $\tilde{\phi}_{l m}(t)$ are defined
in Ref.~\cite{Bohe:2016gbl,Cotesta:2018fcv}.

In the $\SEOB$ model~\cite{Mihaylov:2021bpf}, the complex frequencies
$\sigma_{l m 0}$ are computed by first determining the final mass and spin from
estimates of the initial masses and spins through NR-fitting-formulas~\cite{Taracchini:2013rva,Hofmann:2016yih}
and then converting them to the complex frequencies using BH perturbation
theory-inspiral analytical fits~\cite{Berti:2005ys,Berti:2009kk}.
%
Hence,
%
\begin{subequations}
\begin{align}
\omega_{l m 0}^{\text{GR}} &= \omega_{l m 0}^{\text{GR}}(m_1, m_2, \chi_1, \chi_2)\,,
\\
\tau _{l m 0}^{\text{GR}} &= \tau _{l m 0}^{\text{GR}}(m_1, m_2, \chi_1, \chi_2)\,,
\end{align}
\end{subequations}
%
where $(\omega_{l m 0}^{\text{GR}}, \tau_{l m 0}^{\text{GR}} )$ refer to the
QNM predictions in baseline \SEOB{} model.

@@

Following the ParSpec expansion, our QNM frequencies are expressed through functions,
%
\begin{subequations}
\begin{align}
\omega_k &= \omega_k(m_1, m_2, \chi_1, \chi_2,\ell, \{\delta \omega_k^{(j)}\}),\\
\tau_k   &= \tau _k(m_1, m_2, \chi_1, \chi_2, \ell, \{\delta \tau_k^{(j)}\}),
\end{align}
\end{subequations}
%
where we fixed a certain value of $p$, and the estimates of $(m_1,\, m_2,\, \chi_1,\, \chi_1)$ are used to predict
the final BH mass $M_f$ and spin $\chi_f$ through~\cite{Taracchini:2013rva,Hofmann:2016yih}.
%
Additionally, both $\omega_k$ and $\tau_k$ depend on the coupling constant $\ell$
defined in Eq.~\eqref{eq:def_gamma} as well as the sets, $\{\delta
\omega_k^{(n)}\},$ and $\{\delta \tau_k^{(n)}\}$, where $n$ ranges from zero to
the power of spin in Eqs.~\eqref{eq:kerr_expansion} up to which we include
corrections.
%
For example, if we restrict ourselves to corrections with $n=0$, or
$\chi_f^0$, to the $k=(2,2,0)$ mode only, then we would just have two
additional parameters, $\{\delta \omega_0^{(0)}, \delta \tau_0^{(0)}\}$.
%
In this paper, we will restrict ourselves to corrections to the $n=0$ and
$n=1$ terms, i.e., leading and next-to-leading order corrections to the spin,
$\chi_f$.

@@

% ------------------------------------------------------------------------------------
% HS: we fixed p in our work. We could leave it free in future studies study.
% This is similar to the 'number of dimensions'-like studies. Here it would
% tells us quite generically what is the dimension of the modification to the
% EH-action we include!
% ------------------------------------------------------------------------------------
